Analysing Constraint Grammars with a SAT-solver

### 2015-10-25 paper/lrec/eval_futurework.tex
small stuff

We tested three grammars to find rules that cannot apply: 
Dutch\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}},
with 59 rules; 
Spanish\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}},
with 279 rules; and 
Finnish\footnote{\url{https://github.com/flammie/omorfi/blob/master/src/vislcg3/omorfi.cg3}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule
types introduced in CG-3, and only tested \textsc{remove} and \textsc{select} rules.
The results are shown in table~\ref{table:res}.

@@

\paragraph{Accuracy} 
The test revealed problems in all grammars, and for the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars.

@@

For the Finnish grammar, we did not have time to investigate all the
results. We could verify approximately \nicefrac{1}{3} as true positive and
\nicefrac{1}{3} as false positive.
The verified true positives were all rule-internal conflicts. 
The following rule requires the target to be genitive and unambiguously nominative: 
\begin{itemize}
\item[]\texttt{"oma" SELECT GEN IF [...] (0C NOM) ;}
\end{itemize}
We also detected a number of tagset errors, . 
The grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

@@

\paragraph{Performance} The speed of the check is an obvious issue,
especially for the Finnish grammar. 
As mentioned before, we used 2000 readings extracted from the grammar.
With the full set of 6000 readings from the lexicon, the speed
was even worse: checking the last rule took 30 minutes, as opposed to
3 minutes with 2000 readings. 
%We never finished running the whole rule set with 6000 readings, but
%it would have taken potentially days.
We plan to optimise the code; currently it is just
a naive proof-of-concept implementation.

@@

We will work on the performance, and fix the errors we found in the
evaluation, such as the issue with non-exact position.
We also plan to restrict possible ambiguities in the symbolic words. 
Currently we have only hard-coded some
very basic assumptions, such as ``word cannot be ambiguous between a
boundary and a real word'', but we want to extract these constraints
automatically from a lexicon.
% For instance, the Spanish lexicon contains the entries
% \texttt{casa:casa<n><sg>}  and  \texttt{casa:casar<v><sg><p3>}, thus
% we know that a word can be ambiguous between a noun and a verb.
By adding these restrictions to the symbolic words, we hope to
restrict the search space, and make the SAT problem easier to solve.

>>>>

We tested three grammars to find conflicting rules: 
Dutch\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}},
with 59 rules; 
Spanish\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}},
with 279 rules; and 
Finnish\footnote{\url{https://github.com/flammie/omorfi/blob/master/src/vislcg3/omorfi.cg3}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule
types introduced in CG-3, and only tested \textsc{remove} and \textsc{select} rules.
The results are shown in table~\ref{table:res}.

@@

\paragraph{Accuracy} 
The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars.

@@

For the Finnish grammar, we did not have time to investigate all the
results. We could verify approximately \nicefrac{1}{3} as true positives and
\nicefrac{1}{3} as false positives.
The verified true positives were all rule-internal conflicts. 
For example, the following rule requires the target to be genitive and unambiguously nominative: 
\begin{itemize}
\item[]\texttt{"oma" SELECT GEN IF [...] (0C NOM) ;}
\end{itemize}
We also detected a number of tagset errors;
the grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

@@

\paragraph{Performance} The speed of the check is an obvious issue,
especially for the Finnish grammar. 
As mentioned before, we used 2000 readings extracted from the grammar.
With the full set of 6000 readings from the lexicon, the speed
was even worse: checking the last rule took 30 minutes, as opposed to
3 minutes with 2000 readings. 
%We never finished running the whole rule set with 6000 readings, but
%it would have taken potentially days.
We believe we can still signifcantly improve the performance of the SAT-encoding. Currently it is just a naive proof-of-concept implementation, and we have not tried to optimize for speed at all.

@@

We will work on the performance, and fix the errors we found in the
evaluation, such as the issue with non-exact position.
We also plan to restrict possible ambiguities in the symbolic words. 
Currently we have only hard-coded some
very basic assumptions, such as ``word cannot be ambiguous between a
boundary and a real word'', but we want to extract these constraints
automatically from a lexicon.
% For instance, the Spanish lexicon contains the entries
% \texttt{casa:casa<n><sg>}  and  \texttt{casa:casar<v><sg><p3>}, thus
% we know that a word can be ambiguous between a noun and a verb.
By adding these restrictions to the symbolic words, we hope to
restrict the search space, and make the SAT problem easier to solve, while at the same time generate more realistic example sentences.

### 2015-10-25 paper/lrec/problem_desc.tex
small stuff

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
With our tools, the users can diagnose and improve their grammars.
Some examples of conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

%In addition to analysing the whole grammar,
Our method can be used to
test individual rules. For instance, we can ask the tool to
generate example sequences that trigger rule(s) $R$ but not rule(s) $R'$. 
This can help the grammar writer to see
if they have implemented the rules in the intended way.

>>>>

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
With our tools, the users can find conflicting rules, diagnose problems and improve their grammars.
Some examples of such conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

%In addition to analysing the whole grammar,
%Our method can be used to
%check individual rules. For instance,
We can also ask the tool to
generate example sentences that trigger rule(s) $R$ but not rule(s) $R'$. 
This can help the grammar writer to see
if they have implemented the rules in the intended way.

All of this requires a {\em semantic} rather than a {\em syntactic} analysis.
